---
title: "PM 566 Lab 3"
author: "Jordan Walton"
format:
  html:
    embed-resources: true
---

# **Lab 3 - Exploratory Data Analysis**

```{r}
1 + 1
```


You can add options to executable code like this

```{r}
#| echo: false
2 * 2
```

The `echo: false` option disables the printing of code (only output is displayed).

### **1. Read in the data**

Once you’ve downloaded the file, you can read it into R with:

```{r}
met <- read.csv("~/Downloads/met_all.gz")
```

### **2. Check the dimensions, headers, footers**

**answer: 30 columns, 2377343 rows**

```{r}
dim(met)
head(met)
tail(met)
```

### **3. Take a look at the variables.**

```{r}
str(met)
```

### **4. Take a closer look at the key variables.**

```{r}
table(met$year)
table(met$day)
table(met$hour)
summary(met$temp)
summary(met$elev)
summary(met$wind.sp)
```

It looks like the elevation variable has observations with 9999.0, which is probably an indicator for missing. We should take a deeper look at the data dictionary to confirm. The wind speed variable is OK but there are a lot of missing data.

After checking the data we should make the appropriate modifications. Replace elevations with 9999 as NA.

```{r}
met$elev[met$elev ==9999] <- NA
summary(met$elev)
```

```{r}
met <- met[met$temp > -40, ]
met2 <- met[order(met$temp), ]
head(met2)
```

### **5. Check the data against an external data source**

We removed temperatures colder than -15C. Now, the new data set has a minimum temperature of -3C.

```{r}
met <- met[met$temp>-15,]
met2 <- met [order(met$temp),]
head(met2)
```

### **6. Calculate summary statistics**

```{r}
elev <- met[met$elev==max(met$elev, na.rm=TRUE), ]
summary(elev)
```

```{r}
cor(elev$temp, elev$wind.sp, use="complete")
cor(elev$temp, elev$hour, use="complete")
cor(elev$wind.sp, elev$day, use="complete")
cor(elev$wind.sp, elev$hour, use="complete")
cor(elev$temp, elev$day, use="complete")
```

### **7. Exploratory graphs**

```{r}
hist(met$elev, breaks=100)
hist(met$temp)
hist(met$wind.sp)
```

```{r}
library(leaflet)
leaflet(elev) %>%
  addProviderTiles('OpenStreetMap') %>% 
  addCircles(lat=~lat,lng=~lon, opacity=1, fillOpacity=1, radius=100)
```

```{r}
library(lubridate)
elev$date <- with(elev, ymd_h(paste(year, month, day, hour, sep= ' ')))
summary(elev$date)
elev <- elev[order(is.na(elev$date)), ]
head(elev)
```

```{r}
plot(elev$date, elev$temp, type='l')
plot(elev$date, elev$wind.sp, type='l')
```

Summarize any trends that you see in these time series plots

-   At the end of the graph, there is a slight downward shift in the peaks, which could indicate the beginning of a seasonal cooling trend as summer transitions into fall.

### **8. Ask Questions**

```{r}
hist(elev$wind.sp)
```
